% ============================================================
\section{Background and Problem Formulation}
\label{sec:background}

This section recalls the fault-tolerant architecure our compiler targets, formalises it as a graph, and states the placement problem.
Figure~\ref{fig:overview} previews the pipeline from a circuit to a lattice-surgery schedule.

\input{figures/fig_overview}

\subsection{Surface Codes and Lattice Surgery}

Physical qubits are too noisy to compute on directly, so fault-tolerant quantum
computation (FTQC) encodes each \emph{logical} qubit into many physical qubits
with a quantum error-correcting (QEC) code and operates on the encoded
information while errors are continually detected and corrected. The most widely studied
choice, and the one we target, is the \emph{surface code}, which stores one
logical qubit in a two-dimensional \emph{patch} of physical qubits; enlarging the patch---its code
distance---suppresses the logical error rate
exponentially~\cite{fowler2012surface}. A computation is compiled into the Clifford$+T$ gate set, and each gate becomes an operation on patches.

A logical two-qubit gate is realised by \emph{lattice surgery}. Using the
free ancilla patches in the corridor between two operands, they are \emph{merged} 
into a single patch, which jointly measures
a two-qubit Pauli operator to execute the corresponding gate~\cite{horsman2012lattice,litinski2019game}; the patch is then \emph{split} back into the separate operands. 
The non-Clifford $T$ gate instead consumes a distilled \emph{magic state}, supplied by a magic-state factory. 
It is still executed by the same merge--split operation among 
the operand and the magic state~\cite{bravyi2005magic}.
Both kinds of operation therefore need a free \emph{corridor} of ancilla patches for the
duration of the surgery, and these corridors are the resource the compiler must schedule.

\subsection{Architecture Graph, Safe Passage and Routing}

We model the processor as an \emph{architecture graph} $\A=(V,E)$ over a
rectangular $W\times H$ grid of patches. Each vertex $v\in V$ is a tile with
integer coordinates $(x,y)$; edges $(v,w)\in E \subset V^2$ connect orthogonally adjacent tiles, the
corridors along which surgery can merge patches. A fixed subset
$V_{\mathrm m}\subset V$ hosts magic states, whose number and positions depend on
configuration parameters.

A circuit $\C$ over logical qubits $\Q_L$ is given in the Clifford$+T$ basis. From
it we extract, for each qubit $q$, its $T$-count $t(q)$ (its magic-state demand)
and, for each pair $(q,q')$, the two-qubit interaction count $c(q,q')$; we write
$\bar t,\sigma_t$ and $\bar c,\sigma_c$ for the mean and standard deviation of the
per-qubit $T$- and CNOT-counts. These summary statistics are the only circuit
information the placement uses \todo{optimal config may depend on circuit stats} (Section~\ref{sec:approach}), and the pair counts
define the \emph{CNOT interaction graph} whose density separates sparse from dense circuits.

\begin{definition}[Qubit Mapping]
\label{def:mapping}
    Given a circuit $\C$ over logical qubits $\Q_L$ and architecture $\A=(V,E)$, a \emph{mapping}
    $\M(\A,\C) : \Q_L \to V\setminus V_{\mathrm m}$ is an injective assignment of
    logical qubits to non-magic tiles. A tile in the image of $\M$ or in $V_{\mathrm m}$ 
    is \emph{occupied}, while other tiles are \emph{free}.
\end{definition}

\begin{definition}[Safe Passage]
\label{def:safepassage}
    A (partial) mapping $\M$ satisfies \emph{safe passage} if
    (i) every mapped pair of CNOT partners is joined by a path of free tiles;
    (ii) every mapped qubit $q$ with $t(q)>0$ has a free path from $\M(q)$ to some
    magic tile $m\in V_{\mathrm m}$; and
    (iii) at least one magic tile retains a free neighbour, reserved as an entry
    for the qubits with $T$ demand not yet mapped.
\end{definition}

A mapping that satisfies safe passage is routable. We test this condition in two ways. 
The \emph{connectivity} test checks
(i)--(iii) directly on the free lattice and admits any mapping that preserves
them; it packs patches tightly and fits on smaller grids. The \emph{cube} test is
a stricter, cheaper sufficient condition: it forbids any new patch within the
$3\times3$ neighbourhood of an occupied patch, which needs more area but leaves
more routing options and yields consistently shorter routes in our experiments. We
therefore take \emph{cube} as the default and reserve \emph{connectivity} for
area-constrained lattices.

\begin{definition}[Circuit Layering]
\label{def:layering}
    A \emph{layering} of $\C$ is an ordered partition
    $\Lay(\C)=(L_1,\dots,L_{|\Lay|})$ of its gates into \emph{layers}, where the gates
    of each layer $L_i$ act on pairwise-disjoint qubits and the ordering respects
    every gate dependency. Since dependent gates can never be routed together,
    we can define $\Rmin=\min_{\Lay}|\Lay|$ as the lower-bound for the number of routing steps.
\end{definition}

\begin{definition}[Routing]
\label{def:routing}
    Given an architecture $\A$, a circuit $\C$ and a mapping $\M$, a \emph{routing} 
    is a sequence $\Rout(\A,\C,\M)=(P_1,\dots,P_{|\Rout|})$ where each element
    ---a \emph{routing step}---is a set of vertex-disjoint paths of $\A$. Each gate is 
    realised by a unique path $p\in P_i$ between two tiles. 
    The steps respect the layering $\Lay(\C)$, 
    and the number of \emph{routing steps} is the length $|\Rout|$.
\end{definition}

The router builds $\Rout$ greedily, layer by layer: at each step it serves as many
active gates as it can with disjoint corridors and defers the rest. Serving the
largest such set at once is exactly a maximum vertex-disjoint paths problem on
$\A$, which is $\NP$-hard in
general~\cite{beverland2022surface,siraichi2018qubit}, so practical routers are
heuristic. 

\begin{problem}[Optimally-routable Placement]
    Find a mapping $\M$ that (a) satisfies safe passage and (b) minimises the number of lattice-surgery routing steps $|\Rout|$ required to execute $\C$ on $\A$.
\end{problem}
