% ============================================================
\section{Introduction}
\label{sec:intro}

Fault-tolerant quantum computation protects each logical qubit inside a patch of
a quantum error-correcting code, most commonly the surface code laid out on a
two-dimensional grid of physical qubits~\cite{fowler2012surface}. On such
hardware a logical two-qubit gate is performed via \emph{lattice surgery}, which 
temporarily merges and then splits a strip of
ancilla patches between the two operands~\cite{horsman2012lattice,litinski2019game}. 
The non-Clifford $T$
gate that completes a universal Clifford$+T$ set cannot be applied transversally;
it is teleported by consuming a distilled \emph{magic state}, which must travel
from a magic-state factory to the operand through a free corridor of the
lattice~\cite{bravyi2005magic}. Computation thus becomes a choreography of
patches and of the empty lanes between them.

Compiling a logical circuit onto this fabric splits into two coupled
sub-problems. \emph{Mapping} assigns each logical qubit to a patch; \emph{routing}
then schedules the lattice-surgery operations in time, reserving disjoint ancilla
corridors at every step. Superficially this mirrors the NISQ map-and-route
pipeline of compilers such as SABRE~\cite{li2019sabre} and the Qiskit
transpiler~\cite{qiskit}, but the cost model differs in kind. There are no
$\SWAP{}$ gates and no fixed coupling graph: connectivity is rebuilt from scratch
at each step, and the scarce resource is free \emph{space}---lanes for corridors
and access to magic states. The
initial placement is decisive, because a $T$-heavy qubit stranded far from a
magic state, or a pair of frequent partners scattered across the grid, forces
lattice surgeries to serialise and inflates the routing depth.

We address the placement sub-problem with a \emph{potential-field} heuristic
borrowed from robot motion planning~\cite{khatib1986potential}. Each candidate
tile is scored by a scalar
field obtained by superposing two-dimensional Gaussians: attractive bumps that
pull a $T$-heavy qubit toward magic states and tightly coupled qubits toward
their partners, and repulsive valleys around already-placed patches that keep
routing lanes open. Every qubit is committed to the highest-scoring tile that
passes a \emph{safe-passage} test---a fast reachability check guaranteeing that no
placement traps a patch away from its partners or its magic states---so the
resulting mapping is always routable. The whole pass is deterministic and runs in
low-order polynomial time, in contrast to the search- and solver-based schedulers
that dominate current fault-tolerant toolchains.

Qubit mapping and routing are well studied in the NISQ regime, where heuristics
such as SABRE~\cite{li2019sabre} and the Qiskit transpiler~\cite{qiskit} insert
$\SWAP{}$ gates to satisfy a fixed coupling graph and the underlying allocation
problem is known to be $\NP$-hard~\cite{siraichi2018qubit}; these methods
minimise $\SWAP{}$ count or depth, objectives that do not transfer to a setting
with no $\SWAP{}$s and a per-step reconfigurable topology. On the fault-tolerant
side, Litinski's \emph{game of surface codes}~\cite{litinski2019game} fixes the
lattice-surgery cost model and tile budgets we optimise against, while Beverland
et al.~\cite{beverland2022surface} formalise surface-code routing as an
edge-disjoint-paths problem and compile circuits by solving it. WISQ~\cite{wisq}
is a recent end-to-end fault-tolerant compiler that treats placement and
scheduling jointly with a dependency-aware solver; it produces high-quality
schedules but scales poorly and can time out on larger circuits. Relative to all
of these, our contribution is deliberately lightweight: instead of solving
placement and routing together by search, we compute a cheap, interpretable
placement up front and hand it to a simple downstream router.

\smallskip
\noindent\textbf{Contributions.}
We formulate fault-tolerant initial placement as the maximisation of a Gaussian
potential field over the lattice, sourced by magic states, mapped interaction
partners and already-placed qubits, and we pair it with a safe-passage filter
that certifies routability before any routing is attempted. We give two
weightings of the field---a discrete \emph{coarse} variant and a continuous
\emph{fine} variant that interpolates field strength from per-qubit $T$-count and
per-pair CNOT statistics---and implement everything in an open-source C++20
OpenQASM-to-lattice-surgery compiler~\cite{ftqc}. On \todo{$N$} arithmetic,
algorithmic and state-preparation circuits the field placement reduces routing
steps by \todo{$X$}\% on average over a random baseline and matches
or improves on WISQ (by \todo{$M_{low}-M_{high}\times$}) for \todo{$k$ of $N$} 
circuits, while compiling \todo{$Y$}$\times$ faster (Section~\ref{sec:eval}).
